Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-27 WSSSPE2_report/WSSSPE2_review.tex
Fix a few typos.

We then moved beyond our training discussion to address funding that encourages
sustainable software -- funding of both people and projects that create a true
system to support sustainability. We called this ``institutionalized
serendipity.'' As science is increasingly reliant on software, one participant
observed that ``software development can have much broader impact than
publishing individual research, but it is not viewed that way.'' Because of this
centrality, one participant mentioned that training ought to be called Science
Carpentry rather than Software Carpentry. We felt we were beginning to see
changes in the research community as a result of the NSF's data management plan
requirements. PIs are thinking more about data and some university libraries are
offering data repositories. We wondered if a software management plan might be
effective. ``How might funding programs need to adapt to reward good software
development practices?'' we asked ourselves. We thought about best practices
such as version control, test harnesses, mailing lists, bug tracking, community
contribution, and reuse where appropriate. We thought about measuring success
through usage statistics (downloads, altmetrics). ``Should funders demand
reproducibility?'' we wondered. In order for results to be reproduced, software
would need to be carefully curated.

@@

Possible groups that might give out awards, either under our framework, or more generally,
are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
Society of America, etc. Perhaps this could be a joint technology/science
partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
society] prize.

Some potential criteria for prizes the group suggested are:
community engagement, helping out others;
number of unique contributors;
adding new pieces of functionality to software;
integrating software into broader ecosystem; championing broad principles of
sustainability, open science, open source, etc.;
improving accessibility to software, to scientific software (perhaps championing
inclusiveness or making software accessible?);
documentation;
tutorials;
commits / patching;
leadership transition; and
best contribution by a non-founder.

@@

This group discussed five papers with a wide variety ideas of how to support reproducibility and reuse. It focussed on identifying concrete that the attendees could work together on, which would have a positive effect on the community.

@@

If journals do enforce greater reproducibility constraints, it is important to lower the barriers to reviewers attempting to verifying the correctness of the software used to generate the results. A major issue is that a lot of software only builds on certain systems. Should journals provide more tools / support for reviewers, and if so, what is it? An idea that came from the groups was to define a set of support services that should be available to software paper reviewers. Another was to provide the ability for reviewers to flag the requirement for a `software verification,' similar to ability to flag that a paper needs to be seen by a statistician.

@@

\begin{itemize}
\item Journal Policies
\begin{itemize}
	\item Accessibility of papers
	\begin{itemize}
		\item Open Access
		\item ``Freely'' available
	\end{itemize}
	\item Repositories
	\begin{itemize}
		\item Provides suggestions for recommended repositories
		\item Provides its own repository
	\end{itemize}
	\item Review
	\begin{itemize}
		\item Reviewing software is mandatory
		\begin{itemize}
			\item Must check that software runs
			\item Must check quality of code
			\item Must check performance of code if paper makes claims on relative performance
		\end{itemize}
	\end{itemize}	
	\item Supporting Data
	\begin{itemize}
		\item Must be publicly available
		\item Must be in an Open Access repository
		\item must have a DOI
	\end{itemize}
	\item Article Processing Charge
	\begin{itemize}
		\item APC charge is transparent
		\item APC waiver programme
	\end{itemize}
\end{itemize}

\item Paper Policies
\begin{itemize}
	\item Required Sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are releases
	\begin{itemize}
		\item At no extra cost / at significantly reduced cost
	\end{itemize}
\end{itemize}

\item Software Policies
\begin{itemize}
	\item Software must have a license
	\begin{itemize}
		\item Software must have an open source license
	\end{itemize}
	\item Availability
	\begin{itemize}
		\item Software must be openly available and accessible
	\end{itemize}
	\item Deposit policies
	\begin{itemize}
		\item Software should be in a public repository
		\begin{itemize}
			\item Of particular stature / with preservation plan
		\end{itemize}
		\item Software should have a permanent identifier
		\item Software deposit doesn't count as a prior publication
	\end{itemize}
	\item Runnability and dependencies
	\begin{itemize}
		\item Provide documentation to understand how to run
		\item Provide sample data
		\item Provide dependency information
	\end{itemize}
\end{itemize}
\end{itemize}

@@

This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a``Software Paper''.

@@

Within the group, there were several different paths by which participants had their 
first exposure to code testing:
\begin{itemize}
\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and / or the domain of the participant. 
\item As a way to be confident in other software: When breaking out pieces of software from a larger application, a developer wants to ensure the software works as expected. These types of tests may be ephemeral, though, living long enough to give the author of the test confidence in the code, but not handed on and made available to next user. These experiences can lead to more systematic testing. 
\item Through experience with coding: As you write more software, your confidence in code is reduced and the importance of testing becomes clearer. Attendees joked that they started testing ``when their skepticism/guilt became larger than their arrogance''. 
\item From other projects: When building on top of projects with good testing frameworks, you realize that you should also adopt those good software engineering habits.  
\end{itemize}

@@

Teach testing be beginning with a (smallish) 
piece of code that lacks appropriate tests and develop an exercise that involves 
refactoring the code into a more presentable form through creating / deriving unit 
tests. Students would extract unit-tests through reverse-engineering and / or 
questioning a partner who is a developer/expert.

>>>>

We then moved beyond our training discussion to address funding that encourages
sustainable software -- funding of both people and projects that create a true
system to support sustainability. We called this ``institutionalized
serendipity.'' As science is increasingly reliant on software, one participant
observed that ``software development can have much broader impact than
publishing individual research, but it is not viewed that way.'' Because of this
centrality, one participant mentioned that training ought to be called Science
Carpentry rather than Software Carpentry. We felt we were beginning to see
changes in the research community as a result of the NSF's data management plan
requirements. PIs are thinking more about data and some university libraries are
offering data repositories. We wondered if a software management plan might be
effective. ``How might funding programs need to adapt to reward good software
development practices?'' we asked ourselves. We thought about best practices
such as version control, test harnesses, mailing lists, bug tracking, community
contribution, and reuse where appropriate. We thought about measuring success
through usage statistics (downloads, Altmetrics). ``Should funders demand
reproducibility?'' we wondered. In order for results to be reproduced, software
would need to be carefully curated.

@@

Possible groups that might give out awards, either under our framework, or more generally,
are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
Society of America, etc. Perhaps this could be a joint technology/science
partnership, for example, the [Apache $|$ Mozilla]-[AAAS $|$ disciplinary
society] prize.

Some potential criteria for prizes the group suggested are:
community engagement, helping out others;
number of unique contributors;
adding new pieces of functionality to software;
integrating software into broader ecosystem; championing broad principles of
sustainability, open science, open source, etc.;
improving accessibility to software, to scientific software (perhaps championing
inclusiveness or making software accessible?);
documentation;
tutorials;
commits/patching;
leadership transition; and
best contribution by a non-founder.

@@

This group discussed five papers with a wide variety ideas of how to support reproducibility and reuse. It focused on identifying concrete that the attendees could work together on, which would have a positive effect on the community.

@@

If journals do enforce greater reproducibility constraints, it is important to lower the barriers to reviewers attempting to verifying the correctness of the software used to generate the results. A major issue is that a lot of software only builds on certain systems. Should journals provide more tools/support for reviewers, and if so, what is it? An idea that came from the groups was to define a set of support services that should be available to software paper reviewers. Another was to provide the ability for reviewers to flag the requirement for a `software verification,' similar to ability to flag that a paper needs to be seen by a statistician.

@@

\begin{itemize}
\item Journal Policies
\begin{itemize}
	\item Accessibility of papers
	\begin{itemize}
		\item Open Access
		\item ``Freely'' available
	\end{itemize}
	\item Repositories
	\begin{itemize}
		\item Provides suggestions for recommended repositories
		\item Provides its own repository
	\end{itemize}
	\item Review
	\begin{itemize}
		\item Reviewing software is mandatory
		\begin{itemize}
			\item Must check that software runs
			\item Must check quality of code
			\item Must check performance of code if paper makes claims on relative performance
		\end{itemize}
	\end{itemize}	
	\item Supporting Data
	\begin{itemize}
		\item Must be publicly available
		\item Must be in an Open Access repository
		\item must have a DOI
	\end{itemize}
	\item Article Processing Charge
	\begin{itemize}
		\item APC charge is transparent
		\item APC waiver program
	\end{itemize}
\end{itemize}

\item Paper Policies
\begin{itemize}
	\item Required Sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are releasesat
	\begin{itemize}
		\item At no extra cost/at significantly reduced cost
	\end{itemize}
\end{itemize}

\item Software Policies
\begin{itemize}
	\item Software must have a license
	\begin{itemize}
		\item Software must have an open source license
	\end{itemize}
	\item Availability
	\begin{itemize}
		\item Software must be openly available and accessible
	\end{itemize}
	\item Deposit policies
	\begin{itemize}
		\item Software should be in a public repository
		\begin{itemize}
			\item Of particular stature/with a preservation plan
		\end{itemize}
		\item Software should have a permanent identifier
		\item Software deposit doesn't count as a prior publication
	\end{itemize}
	\item Runnability and dependencies
	\begin{itemize}
		\item Provide documentation to understand how to run
		\item Provide sample data
		\item Provide dependency information
	\end{itemize}
\end{itemize}
\end{itemize}

@@

This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a ``Software Paper''.

@@

Within the group, there were several different paths by which participants had their 
first exposure to code testing:
\begin{itemize}
\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and/or the domain of the participant. 
\item As a way to be confident in other software: When breaking out pieces of software from a larger application, a developer wants to ensure the software works as expected. These types of tests may be ephemeral, though, living long enough to give the author of the test confidence in the code, but not handed on and made available to next user. These experiences can lead to more systematic testing. 
\item Through experience with coding: As you write more software, your confidence in code is reduced and the importance of testing becomes clearer. Attendees joked that they started testing ``when their skepticism/guilt became larger than their arrogance''. 
\item From other projects: When building on top of projects with good testing frameworks, you realize that you should also adopt those good software engineering habits.  
\end{itemize}

@@

Teach testing be beginning with a (smallish) 
piece of code that lacks appropriate tests and develop an exercise that involves 
refactoring the code into a more presentable form through creating/deriving unit 
tests. Students would extract unit-tests through reverse-engineering and/or 
questioning a partner who is a developer/expert.
